### 2019-11-04 submission/response_to_reviewers.md

**1. Is it correct that the study was perfomed on max. 4 mice/treatment group? This might be too low a number, especially in case of the colonization studies to draw conclusions.**

> The reviewer is correct that the number of mice per group was low. We state the exact number of mice per treatment group in Figure 1A (5 mice each for 2 of the treatment groups and 4 mice for the 3rd treatment group). We agree that this is a low number, but would argue that the low microbiome variation across laboratory mice means less mice are needed compared to human epidemiological studies when examining *C. difficile* colonization. Previous *C. difficile* infection mouse model studies from our group and a close collaborator have used 3-5 mice per treatment group (see Koenigsknecht et al. Infection and Immunity 2015 and Theriot et al. Gut Microbes 2011 ) and were still able to draw meaningful conclusions. We have added the following sentence to the discussion to acknowledge the low number of mice used for these experiments as a limitation:
"One limitation of our study is that there were only 4-5 mice per group, which may have limited our ability to identify PPI-induced changes in specific bacteria genera."

**2. The authors should address the fact that they have used only C. difficile str. 630 for their experiments. This is a ribotype 012 strain that is not predominant in the clinical setting. Which ribotypes are typically observed in PPI associated CDI, is there a tendency towards a certain ribotype? Could this have an impact on the interpretation of the data derived from the model described in this manuscript?**

> We thank the reviewers for pointing out that the type of *C. difficile* strain could be an important factor influencing the association between PPIs and CDIs observed in human epidemiological studies. Unfortunately, the studies to date seldom examine the *C. difficile* strains found in PPI users that develop CDIs. The studies that do discuss *C. difficile* strains merely note the type of *C. difficile* frequently found in the region that the study was conducted (Roughead et al. Expert Opinion on Drug Safety 2016; McDonald et al. JAMA Internal Medicine 2015; Chitnis et al. JAMA Internal Medicine 2013; Vardakas et al. International Journal of Infections Diseases 2012; Dalton et al. Alimentary Pharmacology & Therapeutics 2008). Our lab prefers using *C. difficile* 630 for experiments because we are interested in colonization and clearance. *C. difficile* VPI 10463 administration to mice results in a more agressive disease that leads to mortality. Using *C. difficile* 630 after clindamycin administration results in an intermediate phenotype, where the mice are colonized but able to clear infection within 10 days. The intermediate phenotype with *C. difficile* 630 allows us to examine whether the addition of a drug exacerbates or prevents infection and track the mice for a longer period of time. In addition to our lab's prior experience with the *C. difficile* 630 infection mouse model, we also decided to use *C. difficile* 630 because of a recent study that demonstrated three proton pump inhibitors had no effect on *in vitro C. difficile* 630 growth (Maier et al. Nature 2018). We have added the following sentence to the discussion to acknowledge that the type of *C. difficile* strain is another potential contributing factor to the association between PPIs and CDIs seen in human epidemiological studies:
"The type of *C. difficile* strain type could also be an important contributing factor, however our study was limited in that we only tested *C. difficile* 630 (ribotype 012)."

@@

**3. As the manuscript is pretty concise and has in total 4 figures, the supplementary figures can be integrated into the main body.**

> We have submitted this manuscript as an Observation, which is limited to 2 figures total. Additionally, in response to Reviewer #2, we have expanded the size of all the panels in our figures, limiting our ability to include additional panels for the 2 main text figures. Because of the changes to the sizes of the figures, we have opted to keep the supplementary figures as supplemental.

@@

**5. There are some studies that looked into the effect of PPIs on microbiomes in cats and infants, which seem to be in accordance with the findings in this manuscript. Hence, the authors should include these in their discussion:**
**- Schmid SM, et al.,2018**
**- Castellani C, et al., 2017**

> Thank you for the suggestion, we have added the following sentence to the discussion referring to these studies to indicate that other groups have also observed minimal changes in the mammalian microbiome after proton pump inhibitor treatment:
"Interestingly, other groups examining fecal microbiota communities before and after PPI administration to healthy cats and infants with gastroesophageal reflux disease, found PPIs have minimal effects on fecal bacterial community structures, although there were a few significant changes in specific genera (Schmid et al. Frontiers in Veterinary Science 2018; Castellani et al. Frontiers in Cellular and Infection Microbiology 2017)."

**6. It is becoming more and more accepted that Clostridium difficile, due to its biology, is a separate genus, resulting in Clostridioides difficile as the accepted name. The authors should give credit to this and at least state it as Clostridium (Clostridioides) difficile in the text the first time they use it.**

> We appreciate the suggestion and have changed "*Clostridium*" to "*Clostridioides*" in the title, abstract, importance, and introduction sections of our manuscript.

### 2019-11-04 submission/manuscript.Rmd

---
bibliography: references.bib
csl: msphere.csl
fontsize: 11pt
geometry: margin=1.0in
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
---

@@

# The proton pump inhibitor omeprazole does not promote *Clostridium difficile* colonization in a murine model

@@

## Abstract
Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) in humans. We assessed how PPI treatment alters the fecal microbiota and whether treatment promotes CDIs in a mouse model. Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment phase, the day of *C. difficile* challenge, and the following 9 days. We found that mice treated with omeprazole were not colonized by *C. difficile*. When omeprazole treatment was combined with a single clindamycin treatment, one cage of mice remained resistant to *C. difficile* colonization, while the other cage was colonized. Treating mice with only clindamycin followed by challenge resulted in *C. difficile* colonization. 16S rRNA gene sequencing analysis revealed that omeprazole had minimal impact on the structure of the murine microbiota throughout the 16 days of omeprazole exposure. These results suggest omeprazole treatment alone is not sufficient to disrupt microbiota resistance to *C. difficile* infection in mice that are normally resistant in the absence of antibiotic treatment.

## Importance
Antibiotics are the primary risk factor for *Clostridium* (*Clostridioides*) *difficile* infections (CDIs), but other factors may also increase a person's risk. In epidemiological studies, proton pump inhibitor (PPI) use has been associated with CDI incidence and recurrence. PPIs have also been associated with alterations in the human intestinal microbiota in observational and interventional studies. We evaluated the effects of the PPI omeprazole on the structure of the murine intestinal microbiota and its ability to disrupt colonization resistance to *C. difficile*. We found omeprazole treatment had minimal impact on the murine fecal microbiota and did not promote *C. difficile* colonization. Further studies are needed to determine whether other factors contribute to the association between PPIs and CDIs seen in humans or whether aspects of murine physiology may limit its utility to test these types of hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium* (*Clostridioides*) *difficile* infections (CDIs) [@Schubert2015]. It is less clear whether other human medications that impact the microbiota also influence *C. difficile* colonization resistance. Multiple epidemiological studies have suggested an association between proton pump inhibitor (PPI) use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019]. There have also been a number of large cohort studies and interventional clinical trials that demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018]. PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018]. In human fecal samples, PPI use results in increases in *Enterococcaceae*, *Lactobacillaceae*, *Micrococcaceae*, *Staphylococcaceae* and *Streptococcaceae* and decreases in *Ruminococcaceae* [@Imhann2015; @Imhann2017; @Freedberg2015; @maier2018extensive]. Several of these taxa have also been associated with *C. difficile* colonization in humans [@Schuberte01021-14].      
Unfortunately, the studies suggesting a link between PPIs and *C. difficile* were retrospective and did not evaluate changes in the microbiome [@tariq2017association; @nehra2018proton; @Elias2019]. Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use explain the association between PPIs and CDIs. Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].  Here, we evaluated the impact of daily PPI treatment with omeprazole on the murine microbiome and susceptibility to *C. difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C. difficile* to colonize but is mild enough that *C. difficile* is cleared within 10 days [@Jenior2018].

>>>>

---
bibliography: references.bib
csl: msphere.csl
fontsize: 11pt
geometry: margin=1.0in
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
---

@@

# The proton pump inhibitor omeprazole does not promote *Clostridioides difficile* colonization in a murine model

@@

## Abstract
Proton pump inhibitor (PPI) use has been associated with microbiota alterations and susceptibility to *Clostridioides difficile* infections (CDIs) in humans. We assessed how PPI treatment alters the fecal microbiota and whether treatment promotes CDIs in a mouse model. Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole during a 7-day pretreatment phase, the day of *C. difficile* challenge, and the following 9 days. We found that mice treated with omeprazole were not colonized by *C. difficile*. When omeprazole treatment was combined with a single clindamycin treatment, one cage of mice remained resistant to *C. difficile* colonization, while the other cage was colonized. Treating mice with only clindamycin followed by challenge resulted in *C. difficile* colonization. 16S rRNA gene sequencing analysis revealed that omeprazole had minimal impact on the structure of the murine microbiota throughout the 16 days of omeprazole exposure. These results suggest omeprazole treatment alone is not sufficient to disrupt microbiota resistance to *C. difficile* infection in mice that are normally resistant in the absence of antibiotic treatment.

## Importance
Antibiotics are the primary risk factor for *Clostridioides difficile* infections (CDIs), but other factors may also increase a person's risk. In epidemiological studies, proton pump inhibitor (PPI) use has been associated with CDI incidence and recurrence. PPIs have also been associated with alterations in the human intestinal microbiota in observational and interventional studies. We evaluated the effects of the PPI omeprazole on the structure of the murine intestinal microbiota and its ability to disrupt colonization resistance to *C. difficile*. We found omeprazole treatment had minimal impact on the murine fecal microbiota and did not promote *C. difficile* colonization. Further studies are needed to determine whether other factors contribute to the association between PPIs and CDIs seen in humans or whether aspects of murine physiology may limit its utility to test these types of hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridioides difficile* infections (CDIs) [@Schubert2015]. It is less clear whether other human medications that impact the microbiota also influence *C. difficile* colonization resistance. Multiple epidemiological studies have suggested an association between proton pump inhibitor (PPI) use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019]. There have also been a number of large cohort studies and interventional clinical trials that demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018]. PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018]. In human fecal samples, PPI use results in increases in *Enterococcaceae*, *Lactobacillaceae*, *Micrococcaceae*, *Staphylococcaceae* and *Streptococcaceae* and decreases in *Ruminococcaceae* [@Imhann2015; @Imhann2017; @Freedberg2015; @maier2018extensive]. Several of these taxa have also been associated with *C. difficile* colonization in humans [@Schuberte01021-14].      
Unfortunately, the studies suggesting a link between PPIs and *C. difficile* were retrospective and did not evaluate changes in the microbiome [@tariq2017association; @nehra2018proton; @Elias2019]. Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use explain the association between PPIs and CDIs. Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].  Here, we evaluated the impact of daily PPI treatment with omeprazole on the murine microbiome and susceptibility to *C. difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C. difficile* to colonize but is mild enough that *C. difficile* is cleared within 10 days [@Jenior2018].
